---
title: 'Global Conflict Machine Learning'
author: "Timothy Pace"
date: "12/3/2017"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Summary Statistics & Loading of Packages
## Loads Installs & Loads Packages, Reads Conflict Data

```{r}
# Installs required packages
# install.packages("ggplot2")
# install.packages("dplyr")
# install.packages("AUC")
# install.packages("caret")
# install.packages("xtable")

# Loads packages
library("ggplot2")
library("dplyr")
library("AUC")
library("caret")
library("xtable")

options(scipen=999) # Disables scientific notation

conflict_df <- read.csv("ged171.csv") # Loads the data.
```

## Number of All Unique Observations and Features - Conflicts

```{r}
# Rows
nrow(conflict_df)

# Columns
ncol(conflict_df)

# Column Names
colnames(conflict_df)
```

## Summary of Features of Interest - Conflicts.

```{r}
# Selects df of interest.
conflict_df_post_91 <- select(conflict_df, id, conflict_name, year, region, country, country_id, region, where_coordinates, 
                     latitude, longitude, type_of_violence, dyad_name, side_a, deaths_a, 
                     side_b, deaths_b, deaths_civilians, deaths_unknown, low, best, high)

# Removes Cold War Years
conflict_df_post_91 <- conflict_df_post_91[conflict_df_post_91$year >= 1991,]
conflict_df_post_91$type_of_violence <- as.factor(conflict_df_post_91$type_of_violence) # Turns int ToV into factor

# Structure of df.
str(conflict_df_post_91)
```

## Summary Statistics for Observations of Interest - Conflicts.

```{r}
sapply(conflict_df_post_91, summary, na.rm = TRUE, maxsum = 5)
```

# Question #1 Results: What is the predicted trend in the number of global conflicts?

## Linear Regression.

```{r}
# Computes distinct conflict names and distinct years.
conflicts_by_year_df <- conflict_df_post_91 %>% select(conflict_name, year) %>% distinct

# Aggregates count of conflicts by year.
conflicts_by_year_df <- aggregate(conflict_name ~ year, conflicts_by_year_df, length, na.action = na.omit)

mod1 <- lm(conflict_name ~ year, data = conflicts_by_year_df)
summary(mod1)
```

```{r}
plot(conflicts_by_year_df$conflict_name, x=conflicts_by_year_df$year, type="l", 
     main="Distinct Conflicts by Year (1991-2016), Linear Regression", xlab="Year", ylab="Distinct Conflicts")
abline(mod1, col="red")
```

## Polynomial Regression.

```{r}
mod2 <- lm(conflicts_by_year_df$conflict_name ~ poly(conflicts_by_year_df$year, 4))
summary(mod2)
```

```{r}
plot(conflicts_by_year_df$conflict_name, x=conflicts_by_year_df$year, type="l", 
     main="Distinct Conflicts by Year (1991-2016), Polynomial Regression (k = 4)", xlab="Year", ylab="Distinct Conflicts", ylim = c(0,250))
lines(conflicts_by_year_df$year, predict(mod2), col="red")
```

## LOESS Modeling.

```{r}
mod3 <- loess(conflict_name ~ year, data=conflicts_by_year_df, span=0.75, silent = TRUE) # 75% smoothing span
summary(mod3)
```

```{r}
smoothed <- predict(mod3)
plot(conflicts_by_year_df$conflict_name, x=conflicts_by_year_df$year, type="l", main="Distinct Conflicts by Year (1991-2016), LOESS Regression", xlab="Year", ylab="Distinct Conflicts")
lines(smoothed, x=conflicts_by_year_df$year, col="red")
```


# Question 2: What is the predicted trend in the number of total fatalities from global conflict?

## Linear Modeling.

```{r}
total_deaths_df <- aggregate(. ~ year, conflict_df_post_91, sum, na.action = na.omit)

# Best estimate of fatalities vs. year
mod1 <- lm(best ~ year, data = total_deaths_df)
summary(mod1)
```


```{r}
plot(total_deaths_df$best, x=total_deaths_df$year, type="l", 
     main="Global Deaths from Conflict (1991-2016), \nLinear Regression", xlab="Year", ylab="Total Deaths")
abline(mod1, col="red")
```

Removes the Rwandan Genocide from analyses as an outlier.

```{r}
total_deaths_df <- conflict_df_post_91[conflict_df_post_91$conflict_name != 'Government of Rwanda - Civilians',]
total_deaths_no_rg_df <- aggregate(. ~ year, total_deaths_df, sum, na.action = na.omit)
```


```{r}
mod1 <- lm(best ~ year, data = total_deaths_no_rg_df)
summary(mod1)
```

```{r}
plot(total_deaths_no_rg_df$best, x=total_deaths_no_rg_df$year, type="l", 
     main="Global Deaths from Conflict (1991-2016) \nExcluding the Rwandan Genocide (Linear Regression)", xlab="Year", ylab="Total Deaths")
abline(mod1, col="red")
```

## Polynomial Modeling.

```{r}
# Best estimate of fatalities vs. year
mod3 <- lm(total_deaths_no_rg_df$best ~ poly(total_deaths_no_rg_df$year, 4))
summary(mod3)
```

```{r}
plot(total_deaths_no_rg_df$best, x=total_deaths_no_rg_df$year, type="l", 
     main="Global Deaths from Conflict (1991-2016) \nExcluding the Rwandan Genocide (Polynomial Regression, k = 4)", xlab="Year", ylab="Total Deaths", ylim=c(0, 100000))
lines(total_deaths_no_rg_df$year, predict(mod3), col="red")
```

```{r}
mod4 <- loess(best ~ year, data=total_deaths_no_rg_df, span=0.75, silent = TRUE)
summary(mod4)
```

```{r}
smoothed <- predict(mod4)

plot(total_deaths_no_rg_df$best, x=total_deaths_no_rg_df$year, type="l", main="Global Deaths from Conflict (1991-2016) \nExcluding the Rwandan Genocide (LOESS Regression)", xlab="Year", ylab="Total Deaths")
lines(smoothed, x=total_deaths_no_rg_df$year, col="red")
```

# Predicted trend in the number of civilian fatalities from global conflict

## Linear Modeling

```{r}
civilian_deaths_df <- aggregate(. ~ year, conflict_df_post_91, sum, na.action = na.omit)

# Best estimate of fatalities vs. year
mod1 <- lm(deaths_civilians ~ year, data = civilian_deaths_df)
summary(mod1)
```

```{r}
plot(civilian_deaths_df$deaths_civilians, x=civilian_deaths_df$year, type="l", 
     main="Global Civilian Deaths from Conflict (1991-2016), \nLinear Regression", xlab="Year", ylab="Total Deaths")
abline(mod1, col="red")
```

Removes the Rwandan Genocide from analyses as an outlier.

```{r}
civilian_deaths_df <- conflict_df_post_91[conflict_df_post_91$conflict_name != 'Government of Rwanda - Civilians',]
civilian_deaths_no_rg_df <- aggregate(. ~ year, civilian_deaths_df, sum, na.action = na.omit)
```
 
```{r}
mod1 <- lm(deaths_civilians ~ year, data = civilian_deaths_no_rg_df)
summary(mod1)
```

```{r}
plot(civilian_deaths_no_rg_df$deaths_civilians, x=civilian_deaths_no_rg_df$year, type="l", 
     main="Global Civilian Deaths from Conflict (1991-2016) \nExcluding the Rwandan Genocide (Linear Regression)", xlab="Year", ylab="Total Deaths")
abline(mod1, col="red")
```

## Polynomial Modeling.

```{r}
mod3 <- lm(civilian_deaths_no_rg_df$deaths_civilians ~ poly(civilian_deaths_no_rg_df$year, 4))
summary(mod3)
```

```{r}
plot(civilian_deaths_no_rg_df$deaths_civilians, x=civilian_deaths_no_rg_df$year, type="l", 
     main="Global Civilian Deaths from Conflict (1991-2016) \nExcluding the Rwandan Genocide (Polynomial Regression, k = 4)", xlab="Year", ylab="Total Deaths", ylim=c(0, 40000))
lines(civilian_deaths_no_rg_df$year, predict(mod3), col="red")
```

## LOESS Modeling

```{r}
mod4 <- loess(deaths_civilians ~ year, data=civilian_deaths_no_rg_df, span=0.75, silent = TRUE)
summary(mod4)
```

Plotting LOESS regression.

```{r}
smoothed <- predict(mod4)

plot(civilian_deaths_no_rg_df$deaths_civilians, x=civilian_deaths_no_rg_df$year, type="l", main="Global Civilian Deaths from Conflict (1991-2016) \nExcluding the Rwandan Genocide (LOESS Regression)", xlab="Year", ylab="Total Deaths")
lines(smoothed, x=civilian_deaths_no_rg_df$year, col="red")
```

# Factors (e.g. parties involved, region, type of violence, etc.) predicting whether or not there are civilian fatalities in conflict events of violence

## Logistic Regression of Civilian fatalities.

```{r}
# Creates a binary variable for civilian fatalities for logistic regression, and a new DF.
civs_died_lgm_df <- conflict_df_post_91
civs_died_lgm_df$civilians_died <- rep(NA, length(civs_died_lgm_df$deaths_civilians))
civs_died_lgm_df$civilians_died[civs_died_lgm_df$deaths_civilians == 0] <- 0
civs_died_lgm_df$civilians_died[civs_died_lgm_df$deaths_civilians >= 1] <- 1

summary(civs_died_lgm_df$civilians_died)
```

```{r}
summary(civs_died_lgm_df$deaths_civilians)
```

```{r}
str(civs_died_lgm_df)
```


## Runs the initial logistic regression model. 

If quickly re-producing, comment out (takes roughly 30-60 mins).
```{r}
mod1 <- glm(formula = civilians_died ~ deaths_a + side_a + type_of_violence + region,
            family = binomial, data = civs_died_lgm_df, na.action = na.omit)

summary(mod1)
```

## Stepwise Regression.

Performs stepwise regression; If quickly re-producing, comment out (otherwise, leave running overnight or for several hours).
```{r}
mod2 <- step(mod1, trace = 0)
formula(mod2)
```

## Generates the final logistic regression model (takes roughly 30-60 mins).

Civilian fatalities as predicted by "side a" (always a government side or aggressor), deaths on side a, the type of violence (e.g., state-based, non-state, one-sided), and the region. 
```{r}
mod2 <- glm(formula = civilians_died ~ deaths_a + side_a + type_of_violence + region, 
            family = binomial, data = civs_died_lgm_df, na.action = na.omit)
print(summary(mod2))
```

## Quick test case.
```{r}
new_data = data.frame(deaths_a = 0, side_a = 'IS', type_of_violence = "3", region = 'Middle East')
predict(mod2, new_data, type = "response")
```

## Prints the values, sorting by p-values ascending in a table.
```{r, results = 'asis'}
idx <- order(coef(summary(mod2))[,4]) # sorts by p-value ascending
out <- coef(summary(mod2))[idx,] # reorders variablename, coef, SE by p-value ascending
print(xtable(out))
```

## Plots AUC curve.
```{r}
fits <- fitted(mod2)
y <- factor(civs_died_lgm_df$civilians_died)
rr <- roc(fits, y)
plot(rr, main = "ROC Curve")

# AUC Score
auc(rr)
```

## Uses the caret library to perform 10-fold cross validation.
```{r}
# The caret R library requires factors for cross validation logistic regression,
# so we re-create our number of civilian death, dummy variable as a factor.
civs_died_lgm_df$civilians_died_fact <- rep(NA, length(civs_died_lgm_df$deaths_civilians))
civs_died_lgm_df$civilians_died_fact[civs_died_lgm_df$deaths_civilians == 0] <- "No_Civilian_Deaths"
civs_died_lgm_df$civilians_died_fact[civs_died_lgm_df$deaths_civilians >= 1] <- "Civilian_Deaths"

# Create 10 training sets using cross validation.
train_control <- trainControl(method="cv", number=10, classProbs=TRUE, summaryFunction = twoClassSummary)

# Performs cross-validations.
final_mod <- train(civilians_died_fact ~ deaths_a + side_a + type_of_violence + region,
                   data = civs_died_lgm_df, method="glm", family = "binomial",
                   trControl = train_control, metric = "ROC", maximize = TRUE)

# Summarizes the results.
print(final_mod)
```